**Title:** Analyzing and comparing long-term water quality of lakes and reservoirs in Rhode Island and the Northeastern United States using an anomaly approach

### 2020-09-15 manuscript/ecosphere_resubmit/ecosphere_response.Rmd
edits done

- **Response:** We like the reviewers concept of a more holistic way of interpreting trends.  We attempted to do this in our initial assessment of these trends, but we obviously were not successful in doing so.  We now lay out a more explicit approach and describe how we use sign of the slope, magnitude of the slope, relative frequency of high/low years in the beginning and and end of the time frame, and the p-value together to assess for trends.  This is described in the paper in a new section in the methods, *Assessing regressions for trends*.   One of the biggest changes is to suggest values for the slope that would indicate an ecologically meaningful trend.  We describe this in the new section, but in general we are using a trophic state change from oligotrophic to eutrophic over one century to indicate ecologically meaningful change.  For temperature we look at values associated with ecological change as well as policy values (e.g. the Paris Agreement).  While these exact numbers are not universal, we use them as a starting point for assesing the trends.  How we interpreted the trends relative to the additional specific issues raised by the reviewer are detailed below.

@@

- **Response:**  The results for TN in Rhode Island have changed somewhat since we added the additional filtering steps suggested by Reviewer 1 (*1-1* and *1-2*).  This weakened the positive trend somewhat as the slope is smaller and the p-value larger.  Additionally two years, 1995 and 1996, now have insufficient data to be included in the figures.  Using the more explicit definition of a trend we now conclude that TN showed a weak statistically positive trend but not an ecologically important trend over the time frame of a century as the slope is less than the amount needed to cause an trophic state change.  This is reflected in the manuscript.

@@

- **Response:** We applied the same within and across year filters to the LAGOSNE and these results have changed as well.  Given that and our method for interpreting the trends leads to the following:  LAGOSNE is showing weak statistical support for a negative change in nitrogen, with more below average years recently.  The slope is less than the 3.5 we identified, albeit close.  Thus we conclude a weak decrease in TN at the scale of the Region.  This is indicated in the manuscript.

@@

- **Response:** Again, with the changes to the filtering the trend lines have slightly changed.  We concur with Reviewer 1's take that TP is showing a slight increasing trend with some statistical support for a trend, a value greater than the threshold we identified and greater frequency of above average years in recent years.  We have indicated as such in the text.

@@

- **Response:** This is a good suggestion, but we have chosen not to implement it, primarily because of Reveiwer 1's suggestion to have a more holistic approach to addressing the trends.  In developing our approach to assessing the trends we now have ecologically relevant thresholds we can compare to.  Both approaches address the same issue of putting the magnitude of the slopes in context.

@@

- **Response:** NEEDS A RESPONSE!

@@

- **Response:** We agree that this is indeed a timely question and fairly easy to test.  Initially we were going to include this analysis in our revisions and did some preliminary analysis.  There were differences in trends across trophic state with many of the water quality variables and to fully address this would require significant additional analysis and discussion.  After looking at these results and with quite a bit of discussion amongst the co-authors we do still feel that this type of analysis is 1) beyond the scope of this paper and 2) worthy of its own paper.   We have added to the Discussion under *Trends* about this and briefly address preliminary results for differences in chlorophyll trends across just oligotrophic and hypereutrophic lakes.  Our internal conversations are ongoing about completing this analysis for a subsequent paper.

@@

- **Response:** NEEDS A RESPONSE!

@@

- **Response:** NEEDS A RESPONSE!

@@

- **Response:** NEEDS A RESPONSE!

@@

- **Response:** NEEDS A RESPONSE!

@@

- **Response:** NEEDS A RESPONSE!

@@

- **Response:** NEEDS A RESPONSE!

@@

- **Response:** NEEDS A RESPONSE!

>>>>

- **Response:** We like the reviewers concept of a more holistic way of interpreting trends.  We attempted to do this in our initial assessment of these trends, but we obviously were not successful in doing so.  We now lay out a more explicit approach and describe how we use sign of the slope, magnitude of the slope, relative frequency of high/low years in the beginning and and end of the time frame, and the p-value together to assess for trends.  This is described in the paper in a new section in the methods, *Assessing regressions for trends*.   One of the biggest changes is to suggest values for the slope that would indicate an ecologically meaningful trend.  We describe this in the new section, but in general we are using a trophic state change from oligotrophic to eutrophic over one century to indicate ecologically meaningful change.  For temperature we look at values associated with ecological change as well as policy values (e.g. the Paris Agreement).  While these exact numbers are not universal, we use them as a starting point for assessing the trends.  How we interpreted the trends relative to the additional specific issues raised by the reviewer are detailed below.

@@

- **Response:**  The results for TN in Rhode Island have changed somewhat since we added the additional filtering steps suggested by Reviewer 1 (*1-1* and *1-2*) and made changes to use the median instead of the mean in our response to comment *1-6*.  These changes have weakened the positive trend as the slope is smaller and the p-value larger.  Additionally thre years, 1995, 1996, and 1997 now have insufficient data to be included in the figures and a third, 1998, has only three sites.  Using the more explicit definition of a trend we now conclude that TN showed neither a statistically positive trend or an ecologically important trend over the time frame of a century as the slope is less than the amount needed to cause an trophic state change.  This is reflected in the manuscript.

@@

- **Response:** We applied the same within and across year filters to the LAGOSNE, used the median instaed of the mean and these results have changed as well.  Given that and our method for interpreting the trends leads to the following:  LAGOSNE is showing almost no statistical support for a negative change in nitrogen, with more below average years recently.  The slope is much less than the 3.5 we identified.  Thus we conclude no trend in TN at the scale of the Region.  This is indicated in the manuscript.

@@

- **Response:** Again, with the changes to the filtering the trend lines have slightly changed.  We concur with Reviewer 1's take that TP is showing a slight increasing trend with some statistical support for a trend, athough the slope is not greater than the threshold we identified and greater frequency of above average years in recent years.  We have indicated as such in the text.

@@

- **Response:** This is a good suggestion, but we have chosen not to implement it, primarily because of Reviewer 1's suggestion to have a more holistic approach to addressing the trends.  In developing our approach to assessing the trends we now have ecologically relevant thresholds we can compare to.  Both approaches address the same issue of putting the magnitude of the slopes in context.

@@

- **Response:** We were a little unclear about this comment.  Is it referring to coherence over time within a single variable or coherence between the variables?  We agree that for individual variables some years do show greater variability than others.  These increases in the yearly standard deviations are largely the result of a few unusual measurements in those years for those variables.  Essentially these are outliers, but ones that have no obvious reason to remove from the analysis so they are kept in.  Further, there is no pattern to these outliers as they are from different ponds, aren't seen across variables in the same years, and in the case of the LAGOSNE data, they come from different states.  There is only one exception to this and that is the high variability in 2011 for both TN and TP.  That is driven by one sampling event where TN and TP were high for both.  Given the lack of a pattern, there is little additional inference we can draw from this, other than the analysis might be susceptible to outliers.  Given that our original analysis is based on the mean and std. dev. and that those statistics are sensitive to outliers we have decided, thanks to this comment, to change to median and the inter-quartile range, both of which are considerably more robust statistics in the presence of outliers.  With this change the swings in variance are no longer seen as would be expected with statistics that are robust to outliers.  Good catch, by the reviewer.

@@

- **Response:** We agree that this is indeed a timely question and fairly easy to test.  Initially we were going to include this analysis in our revisions and did some preliminary analysis.  There were differences in trends across trophic state with many of the water quality variables and to fully address this would require significant additional analysis and discussion.  After looking at these results and with quite a bit of discussion among the co-authors we do still feel that this type of analysis is 1) beyond the scope of this paper and 2) worthy of its own paper.   We have added to the Discussion under *Trends* about this and briefly address preliminary results for differences in chlorophyll trends across just oligotrophic and hypereutrophic lakes.  Our internal conversations are ongoing about completing this analysis for a subsequent paper.

@@

- **Response:** NEEDS RESPONSE

@@

- **Response:** Added an additional paragraph to the results that compares the two scales and includes a new table summarize the percentage of lakes at both scales in chlorophyll based trophic states.  Another table was added to the trends sections of the results summarizing how we interepreted the trends at each scale.

@@

- **Response:** We changed the wording on this to reflect a focus on the need to know about long term trends.  Also we added three citation that specifically address that sharing results back with volunteers meets their expectations and can motivate continued involvement.

@@

- **Response:** This section was re-written.  We mention nutrient reductions as one possible hypothesis that long-term trends like those seen in RI can inform.

@@

- **Response:** We added a summary table that summarizes average lake area, maximum depth, and average percent forest, ag, and developed for the LAGOSNE and URIWW.  Some additional text is in the methods as well as in the results.  Thanks for the suggestion.  There are a few differences and are a useful addition to the paper.

@@

- **Response:** Thanks again.  We have cleaned up this sentence to better reflect a focus on lengthening growing season and also refer to a source of data (NOAA Climate Time Series) that make this point as well.

@@

- **Response:** In the paragraph about LAGOSNE, we provided some additional clarification.  We use the same selection criteria on both datasets to make sure the time frame is comparable.  The data for LAGOSNE are reported to all be from the surface or epilimnion.  While not a direct match to the depth filter we used for URIWW, they are roughly comparable.  Data for LAGOSNE do not extend beyond 2013, hence the truncation.  This is all reflected in the manuscript.
